RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2014-01-07 doc/bioinformatics2014/document.tex
Bioinformatics2014 folder for rvd27

\documentclass[11pt,reqno]{amsart}
\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{algorithm, algorithmic}

\usepackage{algorithm}
\usepackage{algpseudocode}

\usepackage{graphicx}

\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{enumitem}

\usepackage{setspace}
\doublespacing

\usepackage{natbib}

%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\RR}{I\!\!R} %real numbers
\DeclareMathOperator{\diag}{diag}

\algnewcommand{\Inputs}[1]{%
  \State \textbf{Inputs:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}
\algnewcommand{\Initialize}[1]{%
  \State \textbf{Initialize:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
\author{}
%\date{}                                           % Activate to display a given date or no date

\begin{document}

\begin{abstract}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples. We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
%We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
\end{abstract}

@@

@@

However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples. Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives. Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{li2011statistical,depristo2011framework}. GATK involves more sophisticate pre- and post-processing steps. The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.

Several studies have compared the relative performance of these algorithms.
%TODO: [DISCUSS \citet{Stead:2013fu}, \citet{spencer2013performance}] and others

Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}. These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance. Instead, we focus exclusively on the model structure and statistical inference steps. Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study. We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.

We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}. In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions. We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}. Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.

@@

The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is

\begin{equation}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
\end{equation}

where
\begin{align}
p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
\end{align}

Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
\begin{equation}
p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{equation}

@@

\begin{equation}
\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{equation}

@@

\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
\Repeat
\For {each location j} \Comment{Sample $\mu_j$}
  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
  \State Set $\mu_j$ to the sample median for the T samples

  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
  \EndFor

\EndFor
\Until {sample size sufficient}
\end{algorithmic}
\end{algorithm}

@@

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are
$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
The estimates for the position-level parameters are
$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$
and
$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
The estimates for the genome-level parameters are
$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$
and
$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.

@@

We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$. Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.

@@

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Posterior Density Test}\label{sec:hypothesis_test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
\begin{equation}\label{eqn:bayes_test}
	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
\begin{equation}
	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
\end{equation}

@@

where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. \citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.

@@

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true. Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.

@@

Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}. We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.

@@

The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. A matched normal cells were derived from lymphoblastoid cells from peripheral blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}. The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
% TODO:cite the technical report. %Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx

@@

We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample. To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.

To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g. loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g. homozygous somatic mutation). The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$. The size of the test is $\alpha=0.05$.
%TODO: correct notation from \Pr to summation because we have samples, not distributions.

@@

We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we used the Bayesian test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC}
\end{center}
\end{figure}
% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'

%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
%
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
%\label{fig:MAF}
%\end{center}
%\end{figure}
%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
\label{tbl:comparison_ss}
\end{center}
\end{figure}
% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'

Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance. They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success. VarScan2-somatic is able to call more mixed samples. However, as the read depth increases the specificity degrades. Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below. muTect has good performance across a wide range of MAF levels. But even at the highest depth only has around 0.5 sensitivity for low MAF levels.

@@

Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590. GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels. VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.

\begin{figure}[h]
\begin{center}
\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
\label{tbl:comparison_fdr}
\end{center}
\end{figure}
% from folder 'rvd2\results\2013-09-19_operating_characteristics'

@@

\appendix

>>>>

\documentclass{bioinfo}
\copyrightyear{2005}
\pubyear{2005}
\usepackage{amsmath}
\usepackage{afterpage}
\usepackage{algorithm}
\usepackage{bbm}

%\usepackage{placeins}
\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
\usepackage{xr}
\externaldocument{supplementary_info}
\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}

\newcommand{\RR}{I\!\!R} %real numbers
\DeclareMathOperator{\diag}{diag}
\begin{document}
\firstpage{1}

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

@@

\begin{abstract}

\section{Motivation:}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.

\section{Results:}
We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We identify six mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

\section{Availability:}
Text

\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty}
\end{abstract}

@@

However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.  Samtools/bcftools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}. GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.

Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.  VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}. Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}. With the joint distribution available, one is able to identify locations with dissimilar allele frequencies. muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}. RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{}. However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.

Several studies have compared the relative performance of these algorithms. \citet{spencer2013performance} demonstrated  that VarScan-somatic performed the best comparing with SAMtools, GATK and SPLINTER in detecting minor allele frequencies (MAFs) of 1\% to 8\%, with $ > $500× coverage required for optimal performance. However, \citet{spencer2013performance} also highlighted the fact that VarScan2 yielded more false positives at high read depth. \citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000. We test their performance on much lower MAF variants and across a wider range of coverage depths.

The remainder of this article is organized as follows. In the next section we describe the statistical model structure of our new algorithm, RVD2. Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants. We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions. Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.

@@

The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters can be factorized as

\begin{multline}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) \\ = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
\end{multline}

where

\begin{align}
  \begin{split}
  p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \cdot \\
  &\quad \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
  \end{split}\\
 %
  \begin{split}
  p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \cdot \\
  &\quad \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
  \end{split}\\
 %
  \begin{split}
  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \\
  &\quad \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber\\
  \end{split}\\
\end{align}

Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
\begin{multline}
p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{multline}

@@

\begin{multline}
\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{multline}

@@

  \State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$

  \Repeat

    \For {each location j}

    \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H 
   % \Comment{Sample $\mu_j$}

    \State Set $\mu_j$ to the sample median for the T samples

      \For {each replicate i}

	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
	%\Comment{Sample $\theta_{ji}$}

      \EndFor

    \EndFor

  \Until {sample size sufficient}

  \end{algorithmic}
\end{algorithm}

@@

We present the initial parameter estimates here and provide the derivations in Supplementary Information. The MoM estimate for replicate-level parameters are
$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
The estimates for the position-level parameters are
$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$
and
$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
The estimates for the genome-level parameters are
$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$
and
$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
%While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.

@@

We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \tilde{\mu}_j\cdot (1-\tilde{\mu}_j)$ if $\tilde{\mu}_j \in (10^{-3},1-10^{-3})$ and $\sigma_j = 10^{-4}$ otherwise, where $\tilde{\mu}_j$ is the MoM estimator of $\mu_j$. Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.

@@

%%%%%%%%%%%%%%%%%%
% Posterior Distribution Test
%%%%%%%%%%%%%%%%%%
\subsection{Posterior Distribution Test}\label{sec:hypothesis_test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data. For notational simplicity, we define the random variables associated with these two distributions $\mu_j^{\text{case}}$ and $\mu_j^{\text{control}}$ and the associated samples as $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

A variant is called if $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
\begin{equation}\label{eqn:bayes_test}
	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha,
\end{equation}
where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level. We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the Matthews Correlation Coefficient (MCC) to perfect classification in the ROC curve plot,
\begin{equation}
	\tau^* = \arg\max_\tau \left\{ \text{MCC}(\tau)\right\}.
\end{equation}

@@

where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. \citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available; we choose that value.

@@

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true. We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.

@@

Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield defined MAFs of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}. We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.

@@

The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}. The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. The estimated tumor purity was reported to be 0.8. Matched normal cells were derived from lymphoblastoid cells from peripheral blood. Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}. The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}. The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads. We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
% TODO:cite the technical report. %Not sure the author is correct
% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx

@@

We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample. To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.

To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g. loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g. homozygous somatic mutation). The two hypothesis tests are then $\Pr( \mu_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\Delta} \leq \tau ) > 1-\alpha$. The size of the test is $\alpha=0.05$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
\caption{ROC curve varying read depth showing detection performance.}
\label{fig:ROC}
\end{center}
\end{figure}

% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we used the posterior distribution test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.

@@

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
\label{tbl:comparison_ss}
\end{center}
\end{figure*}

% from folder 'rvd2\results\2013-09-19_operating_characteristics'
% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'

Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance. They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success. GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples. Therefore, those algorithms are expected to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples. VarScan2-somatic is able to call more mixed samples. However, as the read depth increases the specificity degrades. Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below. muTect has good performance across a wide range of MAF levels. But even at the highest depth only has around 0.5 sensitivity for low MAF levels.

@@

Samtools performs well on 100\% MAF sample and performance improves for read depths 3,089 and 30,590. GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels. VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
\label{tbl:comparison_fdr}
\end{center}
\end{figure*}
% from folder 'rvd2\results\2013-09-19_operating_characteristics'
%\afterpage{\clearpage}

@@

The availability of clinical sequence data is increasing as the technical capability to sequence clinical samples at low cost improves. Consequently, we require statistically accurate algorithms that are able to call germline and somatic point mutations in heterogeneous samples with low purity. Such accurate algorithms are a step towards greater access to genomics for clinical diagnostics.
